# **Key Campaign Events and Their Impacts on Polling Performance**
***


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rvest)
library(viridis)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)

```


## Introduction 

## Former Vice President Joe Biden

```{r echo = FALSE, message = FALSE}

poll_data = 
read_csv("./data/cleaned_polls.csv") %>% 
  group_by(end_date, candidate_name) %>% 
  mutate(
    mean_pct = mean(pct)
  )

biden_plot = 
poll_data %>% 
  filter(candidate_name == "Joe Biden") %>% 
  ggplot(aes(x = end_date, y = mean_pct)) +
  geom_point() + 
  geom_line() +
  geom_vline(xintercept = as.numeric(as.Date("2019-03-29")), linetype = 4, color = "blue") +
  geom_vline(xintercept = as.numeric(as.Date("2019-06-28")), linetype = 4, color = "red") +
   geom_vline(xintercept = as.numeric(as.Date("2019-05-01")), linetype = 4, color = "green") +
  labs(
    x = "Date",
    y = "Polling Percent",
    title = "Joe Biden: Polling Performance from January 24th 2019 to November 3, 2019"
  )

ggplotly(biden_plot)
```

#### Lucy Flores’ Essay on Joe Biden’s Inappropriate Touching: March 29, 2019


Lucy Flores a former assemblywoman in the state of Nevada wrote an essay published in The Cut that detailed her experience at a campaign event in 2014 during her run for Lieutenant governor where he smelled her hair and ["proceeded to plant a big slow kiss on the back of my head"]((https://www.thecut.com/2019/03/an-awkward-kiss-changed-how-i-saw-joe-biden.html)). This essay was published less than a month before Vice President Biden announced his campaign for president in April and it was quickly met with a response from the former VP saying [“not once -  never – did I believe I acted inappropriately”](https://www.nytimes.com/2019/03/30/us/politics/joe-biden-lucy-flores-news.html?action=click&module=RelatedCoverage&pgtype=Article&region=Footer) but acknowledged he had made “expressions of affection” over his career and pledged to listen to any accuser or women who shared their experiences. This news was quickly followed by other women speaking out in early April 2019 and sharing similar accounts of inappropriate or uncomfortable touching by the Vice President. 

While the first poll after the publishing of the essay by Flores only showed a 1% decrease from 28% to 27%, the following poll released around the same time as Biden’s official announcement of his bid for the Presidency showed a drastic decline to only 17%. The negative press he received due to these accusations likely contributed to this drop in support in national polls. 


```{r echo = FALSE, message = FALSE}

poll_data %>% 
  ungroup() %>% 
  mutate(poll_date = as.character(end_date)) %>% 
  filter(
    candidate_name == "Joe Biden",
    poll_date %in% c("2019-03-04", "2019-04-15", "2019-04-25")
    )  %>% 
  select(poll_date, mean_pct) %>% 
  arrange(poll_date) %>% 
  knitr::kable()
```

#### Ukraine Conspiracy: Family Conflict of Interest 

On May 1, 2019, the New York Times ran a story on [Joe Biden’s involvement in the firing of Ukraine’s Prosecutor General Viktor Shokin in 2015 that raised questions of a conflict of interest](https://www.nytimes.com/2019/05/01/us/politics/biden-son-ukraine.html). The negotiation that led to Shokin’s firing was touted as a strong performance of Vice President Biden in his foreign policy role while in the White House and the decision to fire Shokin was supported by both the Obama Administration and other Western nations. The prosecutor general was fired for his inability to crack down on the corruption. Hunter Biden, the son of Joe Biden sat on the board of a natural gas company, Burisma,  that according to this New York Times article, was being investigated by Shokin at the time of his firing.

The next poll after this article was published showed the largest gain in polling percentage during the time period covered by these data. After dropping to 17 percent after announcing his candidacy, the next poll ending collection on May 20th, showed an increase of 16 percent to 33 percent overall. While we may have accepted this controversy to negatively impact Joe Biden’s polling performance, this surge in the polls may be due to initial momentum following his bid for presidency or push back against the narrative the Trump Administration had been stressing regarding Biden’s role in Ukraine. 



## Senator Elizabeth Warren

```{r echo = FALSE, message = FALSE}
warren_plot = 
  poll_data %>% 
  filter(candidate_name == "Elizabeth Warren") %>% 
  ggplot(aes(x = end_date, y = mean_pct)) +
  geom_point() + 
  geom_line() +
  geom_vline(xintercept = as.numeric(as.Date("2019-09-16")), linetype = 4, color = "blue") +
  geom_vline(xintercept = as.numeric(as.Date("2019-10-28")), linetype = 4, color = "green") +
   geom_vline(xintercept = as.numeric(as.Date("2019-08-19")), linetype = 4, color = "red") +
   geom_vline(xintercept = as.numeric(as.Date("2019-02-19")), linetype = 4, color = "orange") +
  labs(
    x = "Date",
    y = "Polling Percent",
    title = "Elizabeth Warren: Polling Performance from January 24th 2019 to November 3, 2019"
  )

ggplotly(warren_plot)
```

## Senator Bernie Sanders

```{r echo = FALSE, message = FALSE}

sanders_plot = 
poll_data %>% 
  filter(candidate_name == "Bernie Sanders") %>% 
  ggplot(aes(x = end_date, y = mean_pct)) +
  geom_point() + 
  geom_line() +
  geom_vline(xintercept = as.numeric(as.Date("2019-04-15")), linetype = 4, color = "blue") +
  geom_vline(xintercept = as.numeric(as.Date("2019-10-04")), linetype = 4, color = "green") +
   geom_vline(xintercept = as.numeric(as.Date("2019-10-15")), linetype = 4, color  = "red") +
 geom_vline(xintercept = as.numeric(as.Date("2019-04-25")), linetype = 4, color = "orange") +
  labs(
    x = "Date",
    y = "Polling Percent",
    title = "Bernie Sanders: Polling Performance from January 24th 2019 to November 3, 2019"
  )

ggplotly(sanders_plot)
```

