---
title: "Cleaned Polling Data"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rvest)
library(viridis)
library(plotly)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
	fig.width = 8, 
  fig.height = 6,
  out.width = "90%"
)
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))
```


## Bringing in and cleaning polling data

* restrict to democrats and specific candidates
* restricting grade of the poll (A or A+)
* getting rid of state level polls
* using `end_date` as the polling date used in figures
* renaming candidates to commonly used name

```{r}
all_polls = 
read_csv("./data/president_primary_polls.csv") %>%
   filter(
    is.na(state), 
    party == "DEM",
    fte_grade %in% c("A+", "A"),
    candidate_name %in% c("Amy Klobuchar", "Andrew Yang", "Bernard Sanders", "Elizabeth Warren", "Joseph R. Biden Jr.", "Julián Castro", "Kamala D. Harris", "Pete Buttigieg", "Cory A. Booker", "Tulsi Gabbard")
   ) %>% 
  mutate(
    end_date = as.Date(end_date, format = "%m/%d/%y"),
    candidate_name = recode(candidate_name, "Bernard Sanders" = "Bernie Sanders", "Joseph R. Biden Jr." = "Joe Biden", "Kamala D. Harris" = "Kamala Harris", "Cory A. Booker" = "Cory Booker")
    )
```

## Creating a dataset (`mean_polls`) with a mean polling percent (`mean_pct`) to use in plots 

```{r}
mean_polls = 
  all_polls %>% 
  group_by(end_date, candidate_name) %>% 
  mutate(
    mean_pct = mean(pct)
  )
```

## Plotting average polling percent over time 

```{r}
average_plot = 
mean_polls %>% 

  ggplot(aes(x = end_date, y = mean_pct, color = candidate_name)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Average Polling Over Time By Candidate",
    x = "Time",
    y = "Polling Percent"
  )


ggplotly(average_plot)
```

    
