---
title: "Scraping"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(rvest)
```

Wikipedia frequently updates their data and tables, and this code was used successfully to scrape data on 12/05/2019. Any errors that occur with future use should be able to remedied by updating the problematic column names below. 

# Scraping Data on Debates 
***

#### URL for wikipedia page

```{r}
wiki_debate = read_html("https://en.wikipedia.org/wiki/2020_Democratic_Party_presidential_debates_and_forums")
```

#### Debate 1, Night 1 Scrape

```{r}
debate_time_1_1 = 
wiki_debate %>% 
  html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[7]') %>%
  html_table() 


debate_time_1_1 = debate_time_1_1[[1]] 

debate_time_1_1 = 
  debate_time_1_1 %>% 
  slice(-1, -2, -13) %>% 
  rename(
    "candidate" = "X1",
    "airtime_min" = "X4"
  ) %>% 
  mutate(
    debate = 1,
    night = 1,
    airtime_min = as.numeric(airtime_min)
  ) %>% 
  select(candidate, airtime_min, debate, night)
  

```

#### Debate 1, Night 2 Scrape

```{r}
debate_time_1_2 = 
  wiki_debate %>% 
  html_nodes(xpath = '//*[@id="mw-content-text"]/div/table[8]') %>% 
  html_table()

debate_time_1_2 = debate_time_1_2[[1]] 

debate_time_1_2 = 
  debate_time_1_2 %>% 
  slice(-1, -2, -13) %>% 
  rename(
    "candidate" = "X1",
    "airtime_min" = "X4"
  ) %>% 
  mutate(
    debate = 1,
    night = 2,
    airtime_min = as.numeric(airtime_min)
  ) %>% 
  select(candidate, airtime_min, debate, night)

```

#### Debate 2, Night 1 Scrape

```{r}
debate_time_2_1 = 
  wiki_debate %>% 
  html_node(xpath = '//*[@id="mw-content-text"]/div/table[12]') %>% 
  html_table()

debate_time_2_1 = 
  debate_time_2_1 %>% 
  slice(-1, -2, -13) %>% 
  rename(
    "candidate" = "X1",
    "airtime_min" = "X4"
  ) %>% 
    mutate(
    debate = 2,
    night = 1,
    airtime_min = as.numeric(airtime_min)
  ) %>% 
  select(candidate, airtime_min, debate, night)
```

#### Debate 2, Night 2 Scrape

```{r}
debate_time_2_2 = 
  wiki_debate %>% 
  html_node(xpath = '//*[@id="mw-content-text"]/div/table[13]') %>% 
  html_table()

debate_time_2_2 = 
  debate_time_2_2 %>% 
  slice(-1, -2, -13) %>% 
  rename(
    "candidate" = "X1",
    "airtime_min" = "X4"
  ) %>% 
      mutate(
    debate = 2,
    night = 2,
    airtime_min = as.numeric(airtime_min)
  ) %>% 
  select(candidate, airtime_min, debate, night)

```

#### Debate 3 Scrape

```{r}
debate_time_3 = 
  wiki_debate %>% 
  html_node(xpath = '//*[@id="mw-content-text"]/div/table[17]') %>% 
  html_table()

debate_time_3 = 
  debate_time_3 %>% 
  slice(-11) %>% 
  rename(
    "candidate" = "Candidates",
    "airtime_min" = "Airtime (min.)[111]"
  ) %>% 
     mutate(
    debate = 3,
    night = 1,
    airtime_min = as.numeric(airtime_min)
  ) %>%
  select(candidate, airtime_min, debate, night)
```

#### Debate 4 Scrape

```{r}
debate_time_4 = 
  wiki_debate %>% 
  html_node(xpath = '//*[@id="mw-content-text"]/div/table[20]') %>% 
  html_table()

 debate_time_4 = 
  debate_time_4 %>% 
  slice(-13) %>% 
  rename(
    "candidate" = "Candidates",
    "airtime_min" = "Airtime (min.)[122]"
  ) %>% 
  mutate(
    debate = 4,
    night = 1,
    airtime_min = as.numeric(airtime_min)
  ) %>%
  select(candidate, airtime_min, debate, night)
```

## Binding all speaking time datasets 

```{r}
speaking_time_df = 
  bind_rows(debate_time_1_1, debate_time_1_2, debate_time_2_1, debate_time_2_2, debate_time_3, debate_time_4) %>% 
  filter(
    candidate %in% c("Warren", "Booker", "Klobuchar", "Castro", "Biden", "Sanders", "Buttigieg", "Harris", "Yang", "Gabbard") 
  )
```

